---
title: "Bayesian DiD — Maintenance LLM Tool ROI"
author: "Brolin O'Connell"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(brms)
library(tidybayes)
library(kableExtra)
set.seed(2024)
```

## What Is a Cost Overrun?

A **cost overrun** is the amount by which a job's final cost exceeds its original estimate. If a work order is estimated at \$10,000 and comes in at \$11,800, that is an 18% overrun (ratio = 1.18). Overruns happen for normal reasons — unexpected site conditions, scope changes, parts delays, vendor repricing — but they compound across hundreds of work orders per year. Even a modest systematic reduction in overrun rates recovers real budget.

This analysis asks: **did facilities using the LLM tool experience lower overruns than comparable facilities without it, after controlling for time trends?**

---

## The Model

We estimate the causal effect using Bayesian difference-in-differences:

$$
y_i = \alpha + \beta_1 \, \text{Treated}_i + \beta_2 \, \text{Post}_i + \beta_3 (\text{Treated}_i \times \text{Post}_i) + \epsilon_i
$$

- $y_i = \log(\text{final cost}_i \;/\; \text{estimated cost}_i)$ — the log overrun ratio.
- $\beta_1$ captures pre-existing differences between treated and control sites.
- $\beta_2$ captures the time trend (changes affecting everyone after March).
- $\beta_3$ is the **ATT** (average treatment effect on the treated) — the causal change in overrun attributable to the tool.
- Because the outcome is log-transformed: $\%\;\text{change} = e^{\hat\beta_3} - 1$.

Priors are weakly informative and skeptical of large effects:

$$
\beta_k \sim \mathcal{N}(0,\; 0.5^2), \qquad \sigma \sim \text{Exponential}(2)
$$

Every posterior draw of $\beta_3$ passes through the ROI formula, producing a distribution over ROI rather than a single number:

$$
\text{ROI}^{(s)} = \frac{|e^{\beta_3^{(s)}} - 1| \cdot \bar{C} \cdot N \;-\; \text{Cost}_{\text{tool}}}{\text{Cost}_{\text{tool}}}, \quad s = 1,\dots,S
$$

---

```{r generate-data}
# ── Simulated data ──────────────────────────────────────────────────────────
n_facilities <- 30
n_treated    <- 15
n_wo         <- 600

# ── Enterprise tool costs (year one, 30 locations) ─────────────────────────
license_annual <- 96000    # enterprise SaaS platform
implementation <- 40000    # integration, data migration, config
training       <- 15000    # change management across 15 sites
api_budget     <- 9000     # ongoing LLM API / token spend
total_tool_cost <- license_annual + implementation + training + api_budget
# = $160,000

# Facilities
facilities <- tibble(
  facility = paste0("F", str_pad(1:n_facilities, 2, pad = "0")),
  treated  = c(rep(1L, n_treated), rep(0L, n_facilities - n_treated))
)

# Work orders
wo <- tibble(
  wo_id    = paste0("WO-", str_pad(1:n_wo, 4, pad = "0")),
  facility = sample(facilities$facility, n_wo, replace = TRUE)
) |>
  left_join(facilities, by = "facility") |>
  mutate(
    date = sample(seq(as.Date("2026-01-01"), as.Date("2026-12-31"), by = "day"),
                  n_wo, replace = TRUE),
    type = sample(c("HVAC", "Electrical", "Plumbing", "Elevator"),
                  n_wo, replace = TRUE, prob = c(.35, .30, .20, .15)),

    # Post = after March 1 for treated facilities
    post = if_else(treated == 1L & date >= as.Date("2026-03-01"), 1L, 0L),
    did  = treated * post,

    # Estimated cost by type
    est_cost = case_when(
      type == "HVAC"       ~ rnorm(n(), 7000, 2000),
      type == "Electrical" ~ rnorm(n(), 4000, 1200),
      type == "Plumbing"   ~ rnorm(n(), 2800, 800),
      type == "Elevator"   ~ rnorm(n(), 9000, 3000)
    ) |> pmax(500) |> round(2),

    # Overrun ratio — tool shaves ~3% off overruns (conservative)
    # Control: ~18% average overrun.  Treated-post: ~15% average overrun.
    overrun = case_when(
      did == 1 ~ rnorm(n(), 1.15, 0.14),
      TRUE     ~ rnorm(n(), 1.18, 0.15)
    ) |> pmax(0.85),

    final_cost  = round(est_cost * overrun, 2),
    log_overrun = log(overrun)
  )
```

```{r fit-model, cache=TRUE, results='hide'}
priors <- c(
  prior(normal(0, 0.5),     class = "b"),
  prior(normal(0.15, 0.15), class = "Intercept"),
  prior(exponential(2),     class = "sigma")
)

fit <- brm(
  log_overrun ~ treated + post + did + (1 | facility),
  data   = wo,
  family = gaussian(),
  prior  = priors,
  chains = 4, cores = 4, iter = 3000, warmup = 1000,
  seed   = 42, silent = 2
)
```

## Treatment Effect

```{r att}
draws <- as_draws_df(fit)

att <- draws |>
  transmute(
    att_log = b_did,
    pct     = (exp(att_log) - 1) * 100
  )
```

```{r att-plot, fig.height=3.5}
ggplot(att, aes(pct)) +
  geom_histogram(bins = 55, fill = "#2563EB", alpha = 0.8, color = NA) +
  geom_vline(xintercept = 0, color = "#EF4444", linetype = "dashed", linewidth = 0.9) +
  geom_vline(xintercept = median(att$pct), color = "#F59E0B", linewidth = 1.1) +
  annotate("rect",
    xmin = quantile(att$pct, .025), xmax = quantile(att$pct, .975),
    ymin = -Inf, ymax = Inf, alpha = 0.07, fill = "#2563EB") +
  scale_x_continuous(labels = scales::label_percent(scale = 1)) +
  labs(x = "% Change in Cost Overrun (ATT)", y = "Posterior Draws") +
  theme_minimal(base_size = 12)
```

Median ATT: **`r round(median(att$pct), 1)`%** · 95% CI: [`r round(quantile(att$pct, .025), 1)`%, `r round(quantile(att$pct, .975), 1)`%] · P(effect < 0): **`r round(mean(att$pct < 0) * 100, 1)`%**

---

## ROI

```{r roi}
tp           <- wo |> filter(did == 1)
n_treated_wo <- nrow(tp)
avg_est_cost <- mean(tp$est_cost)

roi <- att |>
  mutate(
    savings = abs(pct / 100) * avg_est_cost * n_treated_wo,
    net     = savings - total_tool_cost,
    roi_pct = net / total_tool_cost * 100
  )
```

```{r roi-table}
tibble(
  Metric = c("Gross Savings (overrun reduction)",
             "Tool Cost (license + setup + training + API)",
             "Net Savings", "ROI %", "P(ROI > 0)"),
  Value  = c(
    scales::dollar(round(median(roi$savings))),
    scales::dollar(total_tool_cost),
    scales::dollar(round(median(roi$net))),
    paste0(round(median(roi$roi_pct), 0), "%"),
    paste0(round(mean(roi$roi_pct > 0) * 100, 0), "%")
  ),
  `95% CI` = c(
    paste0(scales::dollar(round(quantile(roi$savings, .025))), " – ",
           scales::dollar(round(quantile(roi$savings, .975)))),
    "—",
    paste0(scales::dollar(round(quantile(roi$net, .025))), " – ",
           scales::dollar(round(quantile(roi$net, .975)))),
    paste0(round(quantile(roi$roi_pct, .025), 0), "% – ",
           round(quantile(roi$roi_pct, .975), 0), "%"),
    "—"
  )
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(3:5, bold = TRUE, background = "#F0FDF4")
```

```{r roi-plot, fig.height=3.5}
ggplot(roi, aes(roi_pct)) +
  geom_histogram(bins = 55, fill = "#2563EB", alpha = 0.8, color = NA) +
  geom_vline(xintercept = 0, color = "#EF4444", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median(roi$roi_pct), color = "#F59E0B", linewidth = 1.2) +
  annotate("rect",
    xmin = quantile(roi$roi_pct, .025), xmax = quantile(roi$roi_pct, .975),
    ymin = -Inf, ymax = Inf, alpha = 0.07, fill = "#2563EB") +
  annotate("text",
    x = median(roi$roi_pct), y = Inf, vjust = 2, hjust = -0.15,
    size = 4, color = "#F59E0B",
    label = paste0("Median: ", round(median(roi$roi_pct), 0), "%")) +
  scale_x_continuous(labels = scales::label_percent(scale = 1)) +
  labs(x = "ROI (%)", y = "Posterior Draws") +
  theme_minimal(base_size = 12)
```

---

```{r save, include=FALSE}
saveRDS(list(fit = fit, att = att, roi = roi, wo = wo,
             tool_cost = total_tool_cost),
        "bayesian_did_objects.rds")
```
ROI % (101% median): For every dollar invested in the tool, you got about a dollar back in overrun savings (plus your original dollar). The CI running from −91% to +721% reflects the same uncertainty, some draws say you lost most of your investment, others say you made 7x. The wide range is because you're dividing a noisy savings estimate by a fixed $160K fixed more with real data later 